---
title: "Clustering embeddings for unsupervised schema mapping design"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc_depth: 4
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "#>", 
                      collapse = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```


# Setup and data preparation
```{r}
library(cluster)
library(factoextra)
library(dplyr)
```

```{r}
embeddings <- as.matrix(read.csv("data/embeddings.csv", header = FALSE))
unique_values <- readLines("data/unique_values.txt")
rownames(embeddings) <- unique_values
```

```{r}
library(tidyr)
ms <- readr::read_csv("~/OmicsMLRepo/OmicsMLRepoData/inst/extdata/cBioPortal_merging_schema.csv")
expanded_ms <- ms %>%
    select(curated_field, original_field) %>%
    separate_rows(original_field, sep = ";")
```

```{r eval=FALSE, echo=FALSE}
# Assuming your embeddings are in a matrix called 'embeddings'
# with dimensions 673 x 768

# Normalize embeddings for cosine similarity
normalize_vectors <- function(mat) {
  norms <- sqrt(rowSums(mat^2))
  mat / norms
}

normalized_embeddings <- normalize_vectors(embeddings)
```

# Clustering
## K-means with cosine distance
```{r}
# K-Means on normalized embeddings
set.seed(42)
kmeans_result <- kmeans(embeddings, centers = 30, nstart = 25)
clusters_kmeans <- kmeans_result$cluster

# # Find optimal k using elbow method
# wss <- sapply(1:50, function(k) {
#   kmeans(embeddings, centers = k, nstart = 25)$tot.withinss
# })
# plot(1:15, wss, type = "b", xlab = "Number of Clusters", ylab = "WSS",
#      main = "Elbow Method for Optimal k")

# Visualize clusters using PCA (2D projection)
library(ggplot2)
pca_2d <- prcomp(embeddings, center = TRUE, scale. = FALSE)
pca_coords <- as.data.frame(pca_2d$x[, 1:2])
pca_coords$cluster <- as.factor(clusters_kmeans)

ggplot(pca_coords, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2, alpha = 0.6) +
  theme_minimal() +
  labs(title = "K-Means Clustering (PCA Projection)",
       x = "First Principal Component",
       y = "Second Principal Component") +
  theme(legend.position = "right")

# Alternative: Use factoextra for nice visualization
fviz_cluster(kmeans_result, data = embeddings,
             geom = "point",
             ellipse.type = "convex",
             palette = "jco",
             ggtheme = theme_minimal(),
             main = "K-Means Clustering Visualization")

# Silhouette plot
sil <- silhouette(clusters_kmeans, dist(embeddings))
fviz_silhouette(sil)
```

Evaluate clusters
```{r}
# Silhouette score
library(cluster)
sil_score <- silhouette(clusters_kmeans, dist(embeddings))
mean(sil_score[, 3])

# Davies-Bouldin index (lower is better)
library(clusterSim)
index.DB(embeddings, clusters_kmeans)$DB
```

## Fuzzy C-Means Clustering
K-means clustering is an exclusive (hard) clustering method, meaning each data 
point is assigned to one and only one cluster. It does not inherently allow 
elements to overlap. To allow elements to belong to multiple clusters simultaneously, 
we tried Fuzzy C-Means (FCM) clustering, also known as soft clustering or soft K-means. 

FCM with cosine similarity: When vectors are L2-normalized (unit length), the 
Euclidean distance between them is monotonically related to cosine distance. 
Specifically: euclidean_dist² = 2(1 - cosine_similarity)

```{r eval=FALSE}
library(e1071)
library(ggplot2)

## L2 normalize each row
x <- embeddings
x_normalized <- x / sqrt(rowSums(x^2))

## Apply standard FCM
clustObj <- cmeans(x_normalized, 
                   centers = 50,      # number of clusters
                   iter.max = 40,
                   m = 2,            # fuzziness parameter
                   method = "cmeans")

## Access results
clusters <- clustObj$cluster          # hard cluster assignments
membership <- clustObj$membership     # fuzzy membership matrix
centers <- clustObj$centers           # cluster centers

## Extract cluster assignments
cluster_assignments <- as.factor(clustObj$cluster)
length(levels(cluster_assignments)) # the number of final clusters

## Add cluster assignments to the original data frame
x_clustered <- data.frame(x, cluster = cluster_assignments)

## Visualize the clusters using factoextra::fviz_cluster (performs PCA if needed)
fviz_cluster(list(data = x, cluster = cluster_assignments), 
             geom = "point", 
             ellipse.type = "convex", 
             main = "Fuzzy C-Means Clustering Visualization")

## Clusters and their size
table(clustObj$cluster)
```


## vMF mixture models
For true directional/cosine-based clustering, use von Mises-Fisher (vMF) mixture 
models, which are the spherical equivalent of GMM. vMF is the "cosine similarity 
version" of GMM - it models data as lying on a hypersphere and uses cosine similarity naturally.

For sentence embeddings, vMF mixture models are theoretically more appropriate 
than standard GMM because they naturally handle the directional nature of embeddings 
where magnitude is less meaningful than direction.


## Nearest centroid classification with cosine similarity
### Directly from dictionary
```{r eval=FALSE}
source("~/OmicsMLRepo/MetaHarmonizer/EDA/R/concatenate_row_values.R")

dict <- readr::read_csv("~/OmicsMLRepo/OmicsMLRepoData/inst/extdata/cBioPortal_data_dictionary.csv")
context_fields <- c("ColClass", "Description", "AllowedValues")
context <- concatenate_row_values(dict[context_fields])
colname_context <- data.frame(term = dict$ColName, 
                              context = unlist(context))
dim(colname_context)
```

```{r eval=FALSE}
write.csv(colname_context, "data/colname_context.csv", row.names = FALSE)
```

Create target column name embeddings (`target_colname_embeddings_with_context.csv`) with context
(`colname_context.csv`) https://colab.research.google.com/drive/1CdtvIS7HqVIIMKngsGv7I4VMdoIq6WcK#scrollTo=4993b9e6 



### Optimized dictionary embeddings
```{r eval=FALSE}
## SapBERT optimized version of dictionary
url <- "https://docs.google.com/spreadsheets/d/1t2GTvDpgIrR84_ECoft6bQbb2qUr9RoeFtLZWM-ZRDI/edit?usp=sharing"
ss <- googledrive::as_id(url)
dict <- googlesheets4::read_sheet(ss, sheet = "data_dictionary_for_embedding")
write.csv(dict, "data/cbio_dict_for_embed.csv", row.names = FALSE)
```

Confirm that all the token sizes are < 25 for SapBERT
```{r eval=FALSE}
reticulate::source_python('~/OmicsMLRepo/MetaHarmonizer/EDA/Python/token_size.py')
```

Create target column name embeddings (`target_colname_embeddings_with_description.csv`) with context
https://colab.research.google.com/drive/1CdtvIS7HqVIIMKngsGv7I4VMdoIq6WcK#scrollTo=4993b9e6 

```{r}
centroids <- readr::read_csv("data/target_colname_embeddings_with_description.csv", col_names = FALSE) %>% as.matrix
rownames(centroids) <- dict$ColName
centroids[1:4, 1:4]
```

### Optimized dictionary + Clinical-Longformer
```{r clinical_longformer, eval=FALSE}
## Target column description embeddings using Clinical-Longformer
## Mapping accuracy is much worse than SapBERT
centroids <- readr::read_csv("data/target_colname_embeddings_clinical_lf.csv", col_names = TRUE) %>% as.matrix
rownames(centroids) <- dict$ColName
centroids[1:4, 1:4]
```

We continue with the 'optimized-dictionary with SapBERT' (i.e., `target_colname_embeddings_with_description.csv`)

```{r}
source("~/OmicsMLRepo/MetaHarmonizer/EDA/R/centroid_cluster_by_cosine.R")
res <- centroid_cluster_by_cosine(embeddings, centroids)
```

Check the agreement between manually-built harmonization schema and clutersing
based harmonization.

```{r}
source("~/OmicsMLRepo/MetaHarmonizer/EDA/R/check_mapping.R")
mapping_check <- check_mapping(res$similarity, expanded_ms)

round(table(mapping_check$is_mapped_to_curated)/nrow(mapping_check)*100, 2)
```

Top-5 matching
```{r}
top_n <- 5
mapping_check_top_n <- check_mapping(res$similarity, expanded_ms, top_n)
top_n_accuracy <- mapping_check_top_n %>% 
    group_by(original_field) %>% 
    summarise(any_mapped_correctly = any(is_mapped_to_curated))

round(table(top_n_accuracy$any_mapped_correctly)/nrow(top_n_accuracy)*100, 2)
```

Visualize schema mapping accuracy of 673 source columns to 39 target columns:
```{r sm_accuracy_function}
sm_accuracy <- function(mapping, schema, n) {
    tb <- as.data.frame(matrix(NA, nrow = n, ncol = 2))
    colnames(tb) <- c("top_n", "accuracy")
    for (i in seq_len(n)) {
        check <- check_mapping(mapping, schema, top_n = i)
        top_n_accuracy <- check %>% 
            group_by(original_field) %>% 
            summarise(any_mapped_correctly = any(is_mapped_to_curated))

        summary <- table(top_n_accuracy$any_mapped_correctly)
        per_accuracy <- round(summary["TRUE"]/nrow(top_n_accuracy)*100, 2)
        
        tb$top_n[i] <- i
        tb$accuracy[i] <- per_accuracy
    }
    return(tb)
}
```

```{r}
top_n <- 5
top_n_accuracy <- sm_accuracy(res$similarity, expanded_ms, 5)

top_n_accuracy %>%
    ggplot(aes(x = top_n, y = accuracy)) +
    geom_line(color = "blue") +
    labs(title = paste("Schema Mapping accuracy with top", top_n, "nearest embeddings"), 
         x = "Top n embeddings", y = "Accuracy") +
    theme_minimal()
```

### Compare the SchemaMapper result
Schema mapping accuracy of 673 source columns to 39 target columns

```{r}
match <- readr::read_csv("data/test_manual_eval.csv")

top_n_accuracy <- sm_accuracy(match, expanded_ms, 5)
top_n_accuracy %>%
    ggplot(aes(x = top_n, y = accuracy)) +
    geom_line(color = "blue") +
    labs(title = paste("Schema Mapping accuracy with top", top_n, "nearest embeddings"), 
         x = "Top n embeddings", y = "Accuracy") +
    theme_minimal()
```

```{r}
top_n <- 5

## Top-n accuracy from fixed centroid
mapping_check_top_n <- check_mapping(res$similarity, expanded_ms, top_n) 
top_n_accuracy <- mapping_check_top_n %>% 
    group_by(original_field) %>% 
    summarise(mapped_fixed_centroid = any(is_mapped_to_curated))

ind <- which(top_n_accuracy$original_field == "DIAGNOSIS AGE") # manual update of column name
top_n_accuracy$original_field[ind] <- "DIAGNOSIS.AGE"

## Top-n accuracy from SchemaMapper
top_n_accuracy_sm <- check_mapping(match, expanded_ms)

compare <- top_n_accuracy_sm %>%
    select(original_field, centroid, is_mapped_to_curated) %>%
    rename(mapped_schema_mapper = is_mapped_to_curated) %>%
    left_join(top_n_accuracy, by = "original_field")
```

Closer look on the mapping results comparisons
```{r}
## The number of source column that were mapped only by fixed centroid approach
fixed_centroid <- which(compare$mapped_schema_mapper == "FALSE" & 
                            compare$mapped_fixed_centroid == "TRUE")
length(fixed_centroid)

## The number of source column that were failed to be mapped by both approaches
both_failed <- which(compare$mapped_schema_mapper == "FALSE" & 
                         compare$mapped_fixed_centroid == "FALSE")
length(both_failed)

## Actual source column names for the above cases
compare$original_field[fixed_centroid]
compare$original_field[both_failed]
```
We can probably add `TX_` to the rule as 'treatment'.

```{r}
## Easy-to-map columns
both_succeed <- which(compare$mapped_schema_mapper == "TRUE" & 
                         compare$mapped_fixed_centroid == "TRUE")
easy_case <- table(compare$centroid[both_succeed])
# easy_case_df <- stack(easy_case) %>%
#     rename(curated_field = ind, mapped_field_num = values) %>%
#     left_join(., ms[c("curated_field", "original_field_num")], by = "curated_field")

## Hard-to-map columns
rescued <- table(compare$centroid[fixed_centroid]) # mapped by fixed centroid, but not by SchemaMapper
hard_case <- table(compare$centroid[both_failed]) # failed to be mapped by both
```



### Schpperical K-Means (Simpler Alternative)
```{r}
library(skmeans)

# Spherical k-means (uses cosine similarity)
result <- skmeans(X, k = 5, m = 1, method = "genetic")

clusters <- result$cluster
centers <- result$prototypes
```




## Hierarchical clustering with cosine distance
```{r cosine_dist_func}
# Check your data type first
class(embeddings)

# Convert to numeric matrix if needed
embeddings <- as.matrix(embeddings)
embeddings <- apply(embeddings, 2, as.numeric)

# Improved cosine distance function with error checking
cosine_dist <- function(mat) {
  # Ensure it's a numeric matrix
  mat <- as.matrix(mat)
  mat <- apply(mat, 2, as.numeric)
  
  # Calculate norms
  norms <- sqrt(rowSums(mat^2))
  
  # Normalize (avoid division by zero)
  sim <- mat / norms
  
  # Calculate cosine similarity matrix
  sim_matrix <- sim %*% t(sim)
  
  # Convert to distance (1 - similarity)
  dist_matrix <- as.dist(1 - sim_matrix)
  
  return(dist_matrix)
}
```

```{r}
# Calculate cosine distance
dist_matrix <- cosine_dist(embeddings)

# Agglomerative clustering
hc_k <- 30
hc <- hclust(dist_matrix, method = "average")
clusters_hc <- cutree(hc, k = hc_k)

# Visualize dendrogram
plot(hc, labels = FALSE, hang = -1, main = "Hierarchical Clustering Dendrogram")
rect.hclust(hc, k = hc_k, border = "red")

# Visualize clusters in 2D
pca_coords$cluster_hc <- as.factor(clusters_hc)
ggplot(pca_coords, aes(x = PC1, y = PC2, color = cluster_hc)) +
  geom_point(size = 2, alpha = 0.6) +
  theme_minimal() +
  labs(title = "Hierarchical Clustering (PCA Projection)",
       x = "First Principal Component",
       y = "Second Principal Component")

# Fancy dendrogram with factoextra
fviz_dend(hc, k = hc_k, 
          cex = 0.5,
          k_colors = "jco",
          color_labels_by_k = TRUE,
          rect = TRUE,
          main = "Hierarchical Clustering Dendrogram")
```


## HDBSCAN with cosine distance
```{r}
library(dbscan)

# HDBSCAN with cosine distance
dist_matrix <- cosine_dist(embeddings)
hdbscan_result <- hdbscan(dist_matrix, minPts = 4)
clusters_hdbscan <- hdbscan_result$cluster

# View results
table(clusters_hdbscan)  # 0 = noise points

# Plot clusters (excluding noise points)
pca_coords$cluster_hdbscan <- as.factor(clusters_hdbscan)
ggplot(pca_coords, aes(x = PC1, y = PC2, color = cluster_hdbscan, 
                       shape = cluster_hdbscan == 0)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_shape_manual(values = c(19, 4), guide = "none") +  # X for noise
  theme_minimal() +
  labs(title = "HDBSCAN Clustering (PCA Projection)",
       subtitle = "Cluster 0 = Noise Points",
       x = "First Principal Component",
       y = "Second Principal Component")
```


## UMAP Visualization
```{r}
# UMAP (better for visualization)
library(umap)

# UMAP with 2 components
umap_result <- umap(embeddings, n_components = 2)
umap_coords <- as.data.frame(umap_result$layout)
colnames(umap_coords) <- c("UMAP1", "UMAP2")

# Plot K-Means clusters with UMAP
umap_coords$cluster <- as.factor(clusters_kmeans)
ggplot(umap_coords, aes(x = UMAP1, y = UMAP2, color = cluster)) +
  geom_point(size = 2, alpha = 0.6) +
  theme_minimal() +
  labs(title = "K-Means Clustering (UMAP Projection)",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2")

# 3D UMAP visualization (optional)
library(plotly)
umap_3d <- umap(embeddings, n_components = 3)
umap_3d_coords <- as.data.frame(umap_3d$layout)
colnames(umap_3d_coords) <- c("UMAP1", "UMAP2", "UMAP3")
umap_3d_coords$cluster <- as.factor(clusters_kmeans)

plot_ly(umap_3d_coords, 
        x = ~UMAP1, y = ~UMAP2, z = ~UMAP3,
        color = ~cluster,
        type = "scatter3d",
        mode = "markers",
        marker = list(size = 3)) %>%
  layout(title = "3D UMAP Clustering Visualization")
```


# Evaluation
## Comparison plot
```{r}
# Compare all three methods
library(gridExtra)

p1 <- ggplot(pca_coords, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2, alpha = 0.6) +
  theme_minimal() +
  labs(title = "K-Means") +
  theme(legend.position = "none")

p2 <- ggplot(pca_coords, aes(x = PC1, y = PC2, color = cluster_hc)) +
  geom_point(size = 2, alpha = 0.6) +
  theme_minimal() +
  labs(title = "Hierarchical") +
  theme(legend.position = "none")

p3 <- ggplot(pca_coords, aes(x = PC1, y = PC2, color = cluster_hdbscan)) +
  geom_point(size = 2, alpha = 0.6) +
  theme_minimal() +
  labs(title = "HDBSCAN") +
  theme(legend.position = "none")

grid.arrange(p1, p2, p3, ncol = 3, 
             top = "Comparison of Clustering Methods")
```

## Interactive plot with hover information
```{r}
library(plotly)

# Add any metadata you have (e.g., text snippets)
pca_coords$text_label <- paste("Point:", 1:nrow(pca_coords))  # Replace with actual labels

plot_ly(pca_coords, 
        x = ~PC1, 
        y = ~PC2, 
        color = ~cluster,
        text = ~text_label,
        type = "scatter",
        mode = "markers",
        marker = list(size = 8)) %>%
  layout(title = "Interactive Cluster Visualization",
         xaxis = list(title = "PC1"),
         yaxis = list(title = "PC2"))
```


## Evaluation for All Clustering Methods
```{r}
library(cluster)
library(clusterSim)
library(proxy)

# Calculate distance matrix once (using cosine distance)
dist_matrix <- dist(embeddings, method = "cosine")
# Or if using normalized embeddings with Euclidean:
# dist_matrix <- dist(embeddings)

# ----- K-Means Evaluation -----
kmeans_result <- kmeans(embeddings, centers = 10, nstart = 25)
clusters_kmeans <- kmeans_result$cluster

sil_kmeans <- silhouette(clusters_kmeans, dist_matrix)
db_kmeans <- index.DB(embeddings, clusters_kmeans)$DB

cat("K-Means Results:\n")
cat("Silhouette Score:", mean(sil_kmeans[, 3]), "\n")
cat("Davies-Bouldin Index:", db_kmeans, "\n\n")

# ----- Hierarchical Clustering Evaluation -----
hc <- hclust(dist_matrix, method = "average")
clusters_hc <- cutree(hc, k = hc_k)

sil_hc <- silhouette(clusters_hc, dist_matrix)
db_hc <- index.DB(embeddings, clusters_hc)$DB

cat("Hierarchical Clustering Results:\n")
cat("Silhouette Score:", mean(sil_hc[, 3]), "\n")
cat("Davies-Bouldin Index:", db_hc, "\n\n")

# ----- HDBSCAN Evaluation -----
library(dbscan)
hdbscan_result <- hdbscan(dist_matrix, minPts = 5)
clusters_hdbscan <- hdbscan_result$cluster

# IMPORTANT: Exclude noise points (cluster 0) for evaluation
non_noise <- clusters_hdbscan != 0

if (sum(non_noise) > 0 && length(unique(clusters_hdbscan[non_noise])) > 1) {
  # Filter out noise points
  clusters_filtered <- clusters_hdbscan[non_noise]
  embeddings_filtered <- embeddings[non_noise, ]
  dist_matrix_filtered <- dist(embeddings_filtered, method = "cosine")
  
  sil_hdbscan <- silhouette(clusters_filtered, dist_matrix_filtered)
  db_hdbscan <- index.DB(embeddings_filtered, clusters_filtered)$DB
  
  cat("HDBSCAN Results (excluding noise):\n")
  cat("Silhouette Score:", mean(sil_hdbscan[, 3]), "\n")
  cat("Davies-Bouldin Index:", db_hdbscan, "\n")
  cat("Number of noise points:", sum(!non_noise), "\n")
  cat("Number of clusters:", length(unique(clusters_filtered)), "\n\n")
} else {
  cat("HDBSCAN: Not enough valid clusters for evaluation\n\n")
}
```

```{r comprehensive_cluster_evaluation_function}
# Function to evaluate any clustering result
# Function to evaluate any clustering result
evaluate_clustering <- function(clusters, embeddings, dist_matrix = NULL, 
                                method_name = "Clustering") {
  
  # Remove noise points if present (cluster 0 or -1)
  valid_clusters <- clusters[clusters > 0 & clusters != -1]
  valid_indices <- clusters > 0 & clusters != -1
  
  # Check if we have enough clusters
  n_clusters <- length(unique(valid_clusters))
  
  if (n_clusters < 2) {
    cat(method_name, ": Not enough clusters for evaluation\n")
    return(NULL)
  }
  
  # Filter embeddings
  embeddings_filtered <- embeddings[valid_indices, ]
  
  # Calculate distance matrix if not provided
  if (is.null(dist_matrix)) {
    dist_matrix_filtered <- dist(embeddings_filtered, method = "cosine")
  } else {
    # Subset the distance matrix
    dist_matrix_filtered <- as.dist(as.matrix(dist_matrix)[valid_indices, valid_indices])
  }
  
  # Calculate metrics
  sil <- silhouette(valid_clusters, dist_matrix_filtered)
  sil_score <- mean(sil[, 3])
  
  db_score <- index.DB(embeddings_filtered, valid_clusters)$DB
  
  # Calinski-Harabasz index manually (if function not available)
  ch_score <- tryCatch({
    # Try using the package function first
    index.calinski(embeddings_filtered, valid_clusters)
  }, error = function(e) {
    # Calculate manually if function not available
    n <- nrow(embeddings_filtered)
    k <- n_clusters
    
    # Overall mean
    overall_mean <- colMeans(embeddings_filtered)
    
    # Between-cluster sum of squares
    BCSS <- 0
    for (i in unique(valid_clusters)) {
      cluster_points <- embeddings_filtered[valid_clusters == i, , drop = FALSE]
      n_i <- nrow(cluster_points)
      cluster_mean <- colMeans(cluster_points)
      BCSS <- BCSS + n_i * sum((cluster_mean - overall_mean)^2)
    }
    
    # Within-cluster sum of squares
    WCSS <- 0
    for (i in unique(valid_clusters)) {
      cluster_points <- embeddings_filtered[valid_clusters == i, , drop = FALSE]
      cluster_mean <- colMeans(cluster_points)
      WCSS <- WCSS + sum(apply(cluster_points, 1, function(x) sum((x - cluster_mean)^2)))
    }
    
    # Calinski-Harabasz index
    ch <- (BCSS / (k - 1)) / (WCSS / (n - k))
    return(ch)
  })
  
  # Print results
  cat("\n", method_name, "Evaluation:\n", sep = "")
  cat("----------------------------------------\n")
  cat("Number of clusters:", n_clusters, "\n")
  cat("Number of points:", length(valid_clusters), "\n")
  cat("Noise points:", sum(!valid_indices), "\n")
  cat("Silhouette Score:", round(sil_score, 4), 
      "(higher is better, range: -1 to 1)\n")
  cat("Davies-Bouldin Index:", round(db_score, 4), 
      "(lower is better)\n")
  cat("Calinski-Harabasz Index:", round(ch_score, 4), 
      "(higher is better)\n")
  
  # Return results as a list
  return(list(
    n_clusters = n_clusters,
    n_points = length(valid_clusters),
    noise_points = sum(!valid_indices),
    silhouette = sil_score,
    davies_bouldin = db_score,
    calinski_harabasz = ch_score,
    silhouette_object = sil
  ))
}

# Use the function for all methods
kmeans_eval <- evaluate_clustering(clusters_kmeans, embeddings, 
                                   dist_matrix, "K-Means")
hc_eval <- evaluate_clustering(clusters_hc, embeddings, 
                               dist_matrix, "Hierarchical")
hdbscan_eval <- evaluate_clustering(clusters_hdbscan, embeddings, 
                                   dist_matrix, "HDBSCAN")
```

## Compare all methods visually
```{r}
# Create comparison data frame
comparison <- data.frame(
  Method = c("K-Means", "Hierarchical", "HDBSCAN"),
  Silhouette = c(kmeans_eval$silhouette, 
                 hc_eval$silhouette, 
                 hdbscan_eval$silhouette),
  Davies_Bouldin = c(kmeans_eval$davies_bouldin, 
                     hc_eval$davies_bouldin, 
                     hdbscan_eval$davies_bouldin),
  N_Clusters = c(kmeans_eval$n_clusters,
                 hc_eval$n_clusters,
                 hdbscan_eval$n_clusters)
)

print(comparison)

# Plot comparison
library(ggplot2)
library(gridExtra)

p1 <- ggplot(comparison, aes(x = Method, y = Silhouette, fill = Method)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Silhouette Score (Higher is Better)") +
  theme(legend.position = "none")

p2 <- ggplot(comparison, aes(x = Method, y = Davies_Bouldin, fill = Method)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Davies-Bouldin Index (Lower is Better)") +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)
```

```{r}
# Plot silhouette for each method
par(mfrow = c(2, 2))

plot(sil_kmeans, main = "K-Means Silhouette Plot", 
     col = 1:10, border = NA)

plot(sil_hc, main = "Hierarchical Silhouette Plot", 
     col = 1:10, border = NA)

if (!is.null(hdbscan_eval)) {
  plot(hdbscan_eval$silhouette_object, 
       main = "HDBSCAN Silhouette Plot", 
       col = 1:length(unique(clusters_hdbscan[clusters_hdbscan != 0])), 
       border = NA)
}

par(mfrow = c(1, 1))
```

## Compare with manual harmonization
```{r}
# Agglomerative clustering with the same number of clusters as manual (n = 30)
hc <- hclust(dist_matrix, method = "average")
k = 40
sil_cluster <- cutree(hc, k = k)
names(sil_cluster) <- unique_values
sil_cluster <- stack(sil_cluster) %>% rename(cl = values, original_field = ind)

dat <- dplyr::left_join(expanded_ms, sil_cluster, by = "original_field")
```

```{r eval=FALSE}
fname <- paste0("data/cluster_membership/hc_", k,".csv")
write.csv(dat, fname, row.names = FALSE)
```

