---
title: "Untitled"
format: html
editor: visual
---

# Custom inputs

## Training data

All the samples with 661 source columns that were maually harmonized

```{r message=FALSE}
trainDir <- "~/OmicsMLRepo/MetaHarmonizer"
train <- readr::read_csv("~/OmicsMLRepo/MetaHarmonizer/EDA/data/test.csv") # 183,610 x 661
train_gold_standard <- readr::read_csv(file.path(trainDir, "data/schema_mapping_eval/truth.csv"))

head(train)
head(train_gold_standard)
```

```{r train_dict}
## SchemaMapper corpus (i.e., harmonized attributes)
curated_dict_path <- "data/schema/curated_fields.csv"
curated_dict <- readr::read_csv(file.path(trainDir, curated_dict_path))
head(curated_dict)
```

```{r train_alias}
## Mannual mapping information
alias_dict_path <- "data/schema/curated_fields_source_latest_with_flags.csv"
alias_dict <- readr::read_csv(file.path(trainDir, alias_dict_path))
head(alias_dict)
```

## New data

```{r}
test <- readr::read_csv("~/OmicsMLRepo/MetaHarmonizer_Test_Data/source/combined_origin_data.csv") # 4,077 x 246
```

## ALIAS_DICT

**Prompt for Claude Sonnet 4.5**\
The prompt below was used with the cBioPortal data dictionary ([Link](https://github.com/waldronlab/OmicsMLRepoData/blob/master/inst/extdata/cBioPortal_data_dictionary.csv) - commit: 179d8b4)

### Version 1

```         
I'm designing a clinical metadata schema for a cancer genomics database. Here is the list of target/harmonized attributes and their information. Can you create a list of potential/heterogeneous attribute names for each target/harmonized attribute, that are likely to be observed during the clinical metadata collection?
```

This returns `heterogeneous_attribute_mapping.csv` with these columns:

-   `field_name`: The standardized field name\
-   `source`: The alias/source term\
-   `is_numeric_field` (optional): Set to "yes" for numeric fields

```{r}
div1 <- readr::read_csv("data/heterogeneous_attribute_mapping_ver1.csv")
```

### Version 2

```         
I'm designing a clinical metadata schema for a cancer genomics database, MSKCC's cBioPortal. Here is the list of target/harmonized attributes and their information. Can you create a list of potential/heterogeneous attribute names for each target/harmonized attribute that are likely to be observed during the clinical metadata collection? Attributes can vary in their resolution; for example, some columns for harmonized treatment can be individual treatment names with boolean values. Please return the output in a three-column table, including the following columns: - field_name: The standardized field name    - source: The alias/source term    - is_numeric_field (optional): Set to "yes" for numeric fields
```

```{r}
div2 <- readr::read_csv("data/heterogeneous_attribute_mapping_ver2.csv")
```

```{r}
ALIAS_DICT_PATH = "~/OmicsMLRepo/MetaHarmonizer/EDA/data/heterogeneous_attribute_mapping_ver1.csv"
alias_dict <- readr::read_csv(ALIAS_DICT_PATH)
dim(alias_dict)
head(alias_dict, 3)
```

```{r test_sm_alias_from_llm}
CURATED_ALIAS_PATH = "~/OmicsMLRepo/MetaHarmonizer/EDA/data/heterogeneous_attribute_mapping_ver1.csv"
curated_alias <- readr::read_csv(CURATED_ALIAS_PATH)
dim(curated_alias)
head(curated_alias, 3)
```

## CURATED_DICT

```{r test_sm_corpus}
CURATED_DICT_PATH = "~/OmicsMLRepo/MetaHarmonizer_Test_Data/source/curated_dict.csv"
curated_dict <- readr::read_csv(CURATED_DICT_PATH)
dim(curated_dict)
head(curated_dict, 3)
```

# Evaluation

## TRUTH

```{r}
truth <- readr::read_csv("~/OmicsMLRepo/MetaHarmonizer_Test_Data/source/truth_schema_mapper.csv")
```

# Run schema mapper

## TRAIN

```{python train, eval=FALSE}
from src.models.schema_mapper import SchemaMapEngine
file = "~/OmicsMLRepo/MetaHarmonizer/EDA/data/test.csv"
engine = SchemaMapEngine(file, mode="manual", top_k=5)
results = engine.run_schema_mapping()

from evaluation import schema_mapping_evaluation
metrics = schema_mapping_evaluation.compute_accuracy(
     pred_file="~/OmicsMLRepo/MetaHarmonizer/EDA/output/test_auto.csv",
     truth_file="~/OmicsMLRepo/MetaHarmonizer/data/schema_mapping_eval/truth.csv",
     top_k=5,
     save_eval=True,
     out_dir="~/OmicsMLRepo/MetaHarmonizer/EDA/output"
)
```

```{python}
from evaluation import schema_mapping_evaluation
metrics = schema_mapping_evaluation.compute_accuracy(
     pred_file="~/OmicsMLRepo/MetaHarmonizer/EDA/output/train_identified_fields_wide.csv",
     truth_file="~/OmicsMLRepo/MetaHarmonizer/data/schema_mapping_eval/truth.csv",
     top_k=5,
     save_eval=True,
     out_dir="~/OmicsMLRepo/MetaHarmonizer/EDA/output"
)
```


## TEST

```{python test, eval=FALSE}
from src.models.schema_mapper import SchemaMapEngine
file = "/Users/sehyunoh/OmicsMLRepo/MetaHarmonizer_Test_Data/source/combined_origin_data.csv"
engine = SchemaMapEngine(file, mode="auto", top_k=5)
results = engine.run_schema_mapping()

from evaluation import schema_mapping_evaluation
metrics = schema_mapping_evaluation.compute_accuracy(
     pred_file="~/OmicsMLRepo/MetaHarmonizer/EDA/output/combined_origin_data_manual.csv",
     truth_file="~/OmicsMLRepo/MetaHarmonizer_Test_Data/source/truth_schema_mapper.csv",
     top_k=5,
     save_eval=True,
     out_dir="~/OmicsMLRepo/MetaHarmonizer/EDA/output"
)
```

## Visualize

```{r}
eda_output_dir <- "~/OmicsMLRepo/MetaHarmonizer/EDA/output"

train_cbio_alias <- readr::read_csv(file.path(eda_output_dir, "train_cbio_alias_manual_eval.csv"))
colnames(train_cbio_alias)[3] <- "rank"
train_cbio_alias_sub <- train_cbio_alias %>% filter(!is.na(ref_match)) %>% filter(stage != "invalid")

train_llm_alias <- readr::read_csv(file.path(eda_output_dir, "train_llm_alias_manual_eval.csv"))
colnames(train_llm_alias)[3] <- "rank"
train_llm_alias_sub <- train_llm_alias %>% filter(!is.na(ref_match)) %>% filter(stage != "invalid")

train_llm_alias_auto <- readr::read_csv(file.path(eda_output_dir, "train_llm_alias_auto_eval.csv"))
colnames(train_llm_alias_auto)[3] <- "rank"
train_llm_alias_auto_sub <- train_llm_alias_auto %>% filter(!is.na(ref_match)) %>% filter(stage != "invalid")

train_identified <- readr::read_csv(file.path(eda_output_dir, "train_identified_fields_wide_eval.csv"))
colnames(train_identified)[3] <- "rank"
train_identified_sub <- train_identified %>% filter(!is.na(ref_match)) %>% filter(stage != "invalid")

train_no_alias <- readr::read_csv(file.path(eda_output_dir, "train_no_alias_manual_eval.csv"))
colnames(train_no_alias)[3] <- "rank"
train_no_alias_sub <- train_no_alias %>% filter(!is.na(ref_match)) %>% filter(stage != "invalid")

test_cbio_alias <- readr::read_csv(file.path(eda_output_dir, "test_cbio_alias_manual_eval.csv"))
colnames(test_cbio_alias)[3] <- "rank"
test_cbio_alias_sub <- test_cbio_alias %>% filter(!is.na(ref_match)) %>% filter(stage != "invalid")

test_no_alias <- readr::read_csv(file.path(eda_output_dir, "test_no_alias_manual_eval.csv"))
colnames(test_no_alias)[3] <- "rank"
test_no_alias_sub <- test_no_alias %>% filter(!is.na(ref_match)) %>% filter(stage != "invalid")

# test_llm_alias <- readr::read_csv(file.path(eda_output_dir, "test_llm_alias_manual_eval.csv"))
# colnames(test_llm_alias)[3] <- "rank"
# test_llm_alias_sub <- test_llm_alias %>% filter(!is.na(ref_match)) %>% filter(stage != "invalid")
```


```{r}
results_list <- list(
    # "train_cbio_alias" = train_cbio_alias_sub, # 96.2
    "train_llm_alias" = train_llm_alias_sub,
    # "train_llm_alias_auto" = train_llm_alias_auto,
    "train_identified" = train_identified,
    "train_no_alias" = train_no_alias_sub
    # "test_cbio_alias" = test_cbio_alias_sub,
    # "test_no_alias" = test_no_alias_sub
    # "test_llm_alias" = test_llm_alias_sub,
)

plot_multi_accuracy_by_match_position(results_list)
```
