---
title: "Comparative PCA Analysis of Embeddings"
author: "MetaHarmonizer"
date: "`r Sys.Date()`"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## Introduction

This document performs comparative Principal Component Analysis (PCA) on two embedding matrices:

1. **embeddings.csv** - Base embeddings (673 samples)
2. **target_colname_embeddings_with_description.csv** - Target column embeddings with descriptions (39 samples)

The analysis explores the underlying structure, dimensionality, and differences between these two embedding spaces.

```{r setup}
#| label: setup
#| include: true

# Load required libraries
library(ggplot2)
library(factoextra)
library(tidyverse)
library(patchwork)

# Set theme
theme_set(theme_minimal())
```

## Load Data

```{r load-data}
# Load both embedding datasets
embeddings <- read.csv("data/embeddings.csv", header = FALSE)
target_embeddings <- read.csv("data/target_colname_embeddings_with_description.csv", header = FALSE)

cat("Dataset 1 (embeddings.csv):", dim(embeddings)[1], "rows ×", dim(embeddings)[2], "columns\n")
cat("Dataset 2 (target_colname_embeddings_with_description.csv):", dim(target_embeddings)[1], "rows ×", dim(target_embeddings)[2], "columns\n")

# Check if dimensions match
if (ncol(embeddings) != ncol(target_embeddings)) {
  warning("Warning: The two datasets have different numbers of features!")
}
```

```{r perform-pca-1}
# Perform PCA with centering and scaling
pca_1 <- prcomp(embeddings, center = TRUE, scale. = TRUE)

# Calculate variance explained
var_exp_1 <- pca_1$sdev^2 / sum(pca_1$sdev^2)
cum_var_1 <- cumsum(var_exp_1)

# Summary statistics
cat("Dataset 1 (embeddings.csv) - Variance explained by first 10 PCs:\n")
for (i in 1:10) {
  cat(sprintf("PC%d: %.2f%% (Cumulative: %.2f%%)\n", 
              i, var_exp_1[i] * 100, cum_var_1[i] * 100))
}

# Find PCs needed for thresholds
pc_90_1 <- which(cum_var_1 >= 0.90)[1]
pc_95_1 <- which(cum_var_1 >= 0.95)[1]

cat(sprintf("\nPCs needed for 90%% variance: %d\n", pc_90_1))
cat(sprintf("PCs needed for 95%% variance: %d\n", pc_95_1))
```

### Dataset 2: target_colname_embeddings_with_description.csv

```{r pca-dataset2}
#| label: perform-pca-2

# Perform PCA with centering and scaling
pca_2 <- prcomp(target_embeddings, center = TRUE, scale. = TRUE)

# Calculate variance explained
var_exp_2 <- pca_2$sdev^2 / sum(pca_2$sdev^2)
cum_var_2 <- cumsum(var_exp_2)

# Summary statistics
cat("\nDataset 2 (target_colname_embeddings_with_description.csv) - Variance explained by first 10 PCs:\n")
n_show <- min(10, length(var_exp_2))
for (i in 1:n_show) {
  cat(sprintf("PC%d: %.2f%% (Cumulative: %.2f%%)\n", 
              i, var_exp_2[i] * 100, cum_var_2[i] * 100))
}

# Find PCs needed for thresholds
pc_90_2 <- which(cum_var_2 >= 0.90)[1]
pc_95_2 <- which(cum_var_2 >= 0.95)[1]

cat(sprintf("\nPCs needed for 90%% variance: %d\n", pc_90_2))
cat(sprintf("PCs needed for 95%% variance: %d\n", pc_95_2))
```

## Comparative Visualizations

### Scree Plots Comparison

```{r scree-plots-comparison}
#| label: fig-scree-comparison
#| fig-cap: "Comparison of variance explained by principal components"
#| fig-width: 14
#| fig-height: 6

# Create scree plot for dataset 1
p_scree_1 <- fviz_eig(pca_1, 
                      addlabels = TRUE, 
                      ncp = 20,
                      main = "Dataset 1: embeddings.csv",
                      xlab = "Principal Component",
                      ylab = "Percentage of Variance (%)",
                      barfill = "steelblue",
                      barcolor = "steelblue")

# Create scree plot for dataset 2
n_show_2 <- min(20, length(var_exp_2))
p_scree_2 <- fviz_eig(pca_2, 
                      addlabels = TRUE, 
                      ncp = n_show_2,
                      main = "Dataset 2: target_colname_embeddings_with_description.csv",
                      xlab = "Principal Component",
                      ylab = "Percentage of Variance (%)",
                      barfill = "coral",
                      barcolor = "coral")

# Combine plots
p_scree_1 + p_scree_2
```

### Cumulative Variance Comparison

```{r cumulative-variance-comparison}
#| label: fig-cumvar-comparison
#| fig-cap: "Comparison of cumulative variance explained"
#| fig-width: 12
#| fig-height: 6

# Prepare data for both datasets
max_pc <- max(50, length(var_exp_2))
cumvar_df <- data.frame(
  PC = c(1:min(50, length(cum_var_1)), 1:length(cum_var_2)),
  Cumulative = c(cum_var_1[1:min(50, length(cum_var_1))] * 100, cum_var_2 * 100),
  Dataset = c(rep("Dataset 1 (embeddings.csv)", min(50, length(cum_var_1))),
              rep("Dataset 2 (target w/ description)", length(cum_var_2)))
)

ggplot(cumvar_df, aes(x = PC, y = Cumulative, color = Dataset, group = Dataset)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  geom_hline(yintercept = 90, linetype = "dashed", color = "gray40", alpha = 0.5) +
  geom_hline(yintercept = 95, linetype = "dashed", color = "gray60", alpha = 0.5) +
  annotate("text", x = max(cumvar_df$PC) * 0.9, y = 92, label = "90%", color = "gray40", size = 3) +
  annotate("text", x = max(cumvar_df$PC) * 0.9, y = 97, label = "95%", color = "gray60", size = 3) +
  scale_color_manual(values = c("Dataset 1 (embeddings.csv)" = "steelblue", 
                                 "Dataset 2 (target w/ description)" = "coral")) +
  labs(title = "Cumulative Variance Explained - Comparison",
       x = "Principal Component",
       y = "Cumulative Variance (%)",
       color = "Dataset") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        legend.position = "bottom")
```

### PCA Biplots - Dataset 1

#### Dataset 1: PC1 vs PC2

```{r biplot-1-pc1-pc2}
#| label: fig-biplot-1-1-2
#| fig-cap: "Dataset 1 PCA biplot (PC1 vs PC2)"
#| fig-width: 10
#| fig-height: 8

fviz_pca_ind(pca_1,
             axes = c(1, 2),
             geom.ind = "point",
             col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = FALSE,
             title = "Dataset 1: PCA Biplot (PC1 vs PC2)",
             legend.title = "Quality") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

#### Dataset 1: PC1 vs PC3

```{r biplot-1-pc1-pc3}
#| label: fig-biplot-1-1-3
#| fig-cap: "Dataset 1 PCA biplot (PC1 vs PC3)"
#| fig-width: 10
#| fig-height: 8

fviz_pca_ind(pca_1,
             axes = c(1, 3),
             geom.ind = "point",
             col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = FALSE,
             title = "Dataset 1: PCA Biplot (PC1 vs PC3)",
             legend.title = "Quality") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

### PCA Biplots - Dataset 2

#### Dataset 2: PC1 vs PC2

```{r biplot-2-pc1-pc2}
#| label: fig-biplot-2-1-2
#| fig-cap: "Dataset 2 PCA biplot (PC1 vs PC2)"
#| fig-width: 10
#| fig-height: 8

fviz_pca_ind(pca_2,
             axes = c(1, 2),
             geom.ind = "point",
             col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = FALSE,
             title = "Dataset 2: PCA Biplot (PC1 vs PC2)",
             legend.title = "Quality") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

#### Dataset 2: PC1 vs PC3

```{r biplot-2-pc1-pc3}
#| label: fig-biplot-2-1-3
#| fig-cap: "Dataset 2 PCA biplot (PC1 vs PC3)"
#| fig-width: 10
#| fig-height: 8

fviz_pca_ind(pca_2,
             axes = c(1, 3),
             geom.ind = "point",
             col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = FALSE,
             title = "Dataset 2: PCA Biplot (PC1 vs PC3)",
             legend.title = "Quality") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

### PC Score Distributions Comparison

```{r pc-distributions-comparison}
#| label: fig-distributions-comparison
#| fig-cap: "Comparison of PC score distributions for the first 4 principal components"
#| fig-width: 14
#| fig-height: 10

# Create dataframe for both datasets
pc_scores_1 <- data.frame(pca_1$x[, 1:4])
pc_scores_1$Dataset <- "Dataset 1"
pc_scores_1$ID <- 1:nrow(pc_scores_1)

pc_scores_2 <- data.frame(pca_2$x[, 1:4])
pc_scores_2$Dataset <- "Dataset 2"
pc_scores_2$ID <- 1:nrow(pc_scores_2)

# Combine and reshape to long format
pc_scores_combined <- rbind(pc_scores_1, pc_scores_2) %>%
  pivot_longer(cols = starts_with("PC"), 
               names_to = "Component", 
               values_to = "Score")

# Plot distributions with overlay
ggplot(pc_scores_combined, aes(x = Score, fill = Dataset, color = Dataset)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  facet_wrap(~ Component, scales = "free", nrow = 2) +
  labs(title = "Comparison of PC Score Distributions",
       x = "PC Score",
       y = "Density") +
  scale_fill_manual(values = c("Dataset 1" = "steelblue", "Dataset 2" = "coral")) +
  scale_color_manual(values = c("Dataset 1" = "steelblue", "Dataset 2" = "coral")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        legend.position = "bottom")
```

### Side-by-Side PC Scatter Plots

```{r combined-scatter-comparison}
#| label: fig-combined-comparison
#| fig-cap: "Side-by-side comparison of PC1 vs PC2 for both datasets"
#| fig-width: 14
#| fig-height: 6

# Prepare data for both datasets
pc_df_1 <- as.data.frame(pca_1$x[, 1:2])
pc_df_1$Dataset <- "Dataset 1"

pc_df_2 <- as.data.frame(pca_2$x[, 1:2])
pc_df_2$Dataset <- "Dataset 2"

# Plot for Dataset 1
p_scatter_1 <- ggplot(pc_df_1, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.6, color = "steelblue", size = 2.5) +
  labs(title = "Dataset 1: embeddings.csv",
       x = paste0("PC1 (", sprintf("%.1f", var_exp_1[1] * 100), "%)"),
       y = paste0("PC2 (", sprintf("%.1f", var_exp_1[2] * 100), "%)")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12))

# Plot for Dataset 2
p_scatter_2 <- ggplot(pc_df_2, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.6, color = "coral", size = 2.5) +
  labs(title = "Dataset 2: target w/ description",
       x = paste0("PC1 (", sprintf("%.1f", var_exp_2[1] * 100), "%)"),
       y = paste0("PC2 (", sprintf("%.1f", var_exp_2[2] * 100), "%)")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12))

# Combine plots
p_scatter_1 + p_scatter_2 +
  plot_annotation(title = "PC Space Comparison: Both Datasets",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16)))
```

### Variance Explained Comparison Table

```{r variance-table}
#| label: tbl-variance
#| tbl-cap: "Comparison of variance explained by first 10 PCs"

# Create comparison table
n_comp <- min(10, min(length(var_exp_1), length(var_exp_2)))
variance_comparison <- data.frame(
  PC = 1:n_comp,
  Dataset1_Individual = sprintf("%.2f%%", var_exp_1[1:n_comp] * 100),
  Dataset1_Cumulative = sprintf("%.2f%%", cum_var_1[1:n_comp] * 100),
  Dataset2_Individual = sprintf("%.2f%%", var_exp_2[1:n_comp] * 100),
  Dataset2_Cumulative = sprintf("%.2f%%", cum_var_2[1:n_comp] * 100)
)

knitr::kable(variance_comparison,
             col.names = c("PC", 
                          "Dataset 1 (%)", 
                          "Dataset 1 Cumulative (%)",
                          "Dataset 2 (%)",
                          "Dataset 2 Cumulative (%)"),
             align = c("c", "r", "r", "r", "r"))
```

## Summary

::: {.callout-note}
## Key Findings

### Dataset 1: embeddings.csv
- **Total observations**: `r nrow(embeddings)` samples
- **Total dimensions**: `r ncol(embeddings)` features
- **PC1 variance**: `r sprintf("%.2f%%", var_exp_1[1] * 100)`
- **PC2 variance**: `r sprintf("%.2f%%", var_exp_1[2] * 100)`
- **PC3 variance**: `r sprintf("%.2f%%", var_exp_1[3] * 100)`
- **Top 3 PCs capture**: `r sprintf("%.2f%%", cum_var_1[3] * 100)` of total variance
- **PCs for 90% variance**: `r pc_90_1` components
- **PCs for 95% variance**: `r pc_95_1` components

### Dataset 2: target_colname_embeddings_with_description.csv
- **Total observations**: `r nrow(target_embeddings)` samples
- **Total dimensions**: `r ncol(target_embeddings)` features
- **PC1 variance**: `r sprintf("%.2f%%", var_exp_2[1] * 100)`
- **PC2 variance**: `r sprintf("%.2f%%", var_exp_2[2] * 100)`
- **PC3 variance**: `r sprintf("%.2f%%", var_exp_2[3] * 100)`
- **Top 3 PCs capture**: `r sprintf("%.2f%%", cum_var_2[3] * 100)` of total variance
- **PCs for 90% variance**: `r pc_90_2` components
- **PCs for 95% variance**: `r pc_95_2` components
:::


## Session Info

```{r session-info}
#| label: session-info

sessionInfo()
```
